Enabling Progressive Server-Side Rendering for Traditional Web Template Engines with Java Virtual Threads

### 2025-08-07 docs/article/software-3702556.tex

\begin{table}[H]
  \small
    \caption{
    \hl{Comparison} %MDPI: 1. Please confirm whether the different font colors are retained. If so, please confirm whether additional explanation is needed. 2. Please confirm whether √ and × require additional explanation. 3. Please confirm whether the bold text in the table can be deleted
 of web template technologies in the Java ecosystem.
$^{(1)}$ Enforced via annotation processor that checks templates against typed interfaces at compile time.
$^{(2)}$ Compile-time expression validation available via \texttt{@CheckedTemplate} and build-time metadata in Quarkus.
$^{(3)}$ \texttt{Publisher} of reactive streams~\cite{ReactiveStreams}. Limited to a single model per web template.
$^{(4)}$ \texttt{kotlin.sequences.Sequence}
$^{(5)}$ {Non-safety} for HTML attributes.}

@@

  % Subfloat 1
\begin{minipage}[c]{0.33\linewidth}
  \centering
  \begin{lstlisting}[
  language=java,
  basicstyle=\scriptsize\ttfamily,
  numbers=none
]
html();
 head();
  title();text("JT");end();
 end();
 body();
  p();
   text("Hi JATL");
  end();
 end();
end();
  \end{lstlisting}
    \caption*{\centering(\textbf{a}) Function sequence}
  \end{minipage}
  \hfil
  % Subfloat 2
  \begin{minipage}[c]{0.23\linewidth}
    \centering
    \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html(
 head(
  title("ST")
 ),
 body(
  p("Hi ScalaTags")
 )
);
  \end{lstlisting}
    \caption*{\centering(\textbf{b}) Nested function}
  \end{minipage}
  \hfil
  % Subfloat 3
  \begin{minipage}[c]{0.32\linewidth}
    \centering
    \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html()
 .head()
  .title().text("HF").__()
 .__()
 .body()
  .p()
   .text("Hi HtmlFlow")
  .__()
 .__()
.__();
  \end{lstlisting}
    \caption*{\centering(\textbf{c}) Method chaining}
  \end{minipage}
\vspace*{6pt}
  \caption{\hl{Utilizing} %MDPI: Please note that changes to the position/size of figures or tables may occur during the production stage.
 DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
  \label{fig:dsl-idioms}
\end{figure}

@@

All benchmarks were conducted on a GitHub-hosted virtual machine under \hl{GitHub} %MDPI: Please state which version of the software was used.
Actions, running \hl{Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
(x86\_64 architecture)} %MDPI: Please state the name of the manufacturer, city, and country from where the equipment was sourced.
 operating at 3.24 GHz, configured with two CPU cores and two
threads per core, and 7.8 GB of available RAM. The system utilizes a 75 GB SSD
with ext4 file system, achieving 1.5 GB/s write throughput in basic I/O tests.
Network connectivity is provided through a 1500 MTU Ethernet interface. All
tests were conducted on OpenJDK 21 (Corretto build) with the G1 garbage
collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
     -Xmx16g}. Each test was executed five times per configuration to ensure result
reliability.

@@

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring WebFlux with \texttt{Presentations} class.}\label{fig:presentations-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:presentations-webflux-jmeter} show that when
using blocking template engines with a separate coroutine dispatcher, the
engines are unable to scale effectively beyond four concurrent users. In contrast,
\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4487 %MDPI: Commas are only used for numbers with five or more digits. We have removed them in four-digit numbers. Please confirm.
requests per second. When blocking approaches are executed in the context of
virtual threads---thus enabling non-blocking I/O---the engines scale up to 64
concurrent users, with JStachio using virtual threads reaching 3514 requests
per second. The Thymeleaf implementation using the reactive View Resolver
driver scales up to 32 concurrent users, achieving a lower maximum throughput
of 2559 requests per second.

@@

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring MVC with \texttt{Presentations} class.}\label{fig:presentations-springmvc-jmeter}
\end{figure}

@@

\vspace{-3pt}
\begin{figure}[H]
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Throughput %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Quarkus with \texttt{Presentations} class.}\label{fig:presentations-quarkus-jmeter}
\end{figure}

@@

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring WebFlux with \texttt{Stocks} class.}\label{fig:stocks-webflux-jmeter}
\end{figure}

@@

\begin{figure}[H]
\vspace{-3pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring MVC with \texttt{Stocks} class.}\label{fig:stocks-springmvc-jmeter}
\end{figure}
\vspace{-6pt}
\begin{figure}[H]

     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Quarkus with \texttt{Stocks}.}\label{fig:stocks-quarkus-jmeter}
\end{figure}

@@

\bibitem[Carvalho(2025)]{wise2024pssr}
Carvalho, F.M.
\newblock Progressive Server-Side Rendering with Suspendable Web Templates.
\newblock In \emph{Web Information Systems Engineering---WISE
  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; Springer%MDPI: newly added, please confirm it
: Singapore,  2025; pp. 458--473.

@@

\bibitem[Carvalho(2017)]{htmlflow}
Carvalho, F.M.
\newblock HtmlFlow Java DSL to Write Typesafe HTML.
\newblock Technical Report, 2017. Available online: \url{https://htmlflow.org/} accessed on 6 August 2025. %MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.

@@

\bibitem[Carvalho et~al.(2020)Carvalho, Duarte, and Gouesse]{carvalho2020}
Carvalho, F.M.; Duarte, L.; Gouesse, J.
\newblock Text Web Templates Considered Harmful.
\newblock In {\em Lecture Notes in Business Information Processing}; \hl{{2020};} %MDPI: We are sorry but we could not find the required information about this entry. Please add the name of the publisher and their location.
 pp. 69--95.

@@

\bibitem[Sussman and Steele(1975)]{scheme}
Sussman, G.; Steele, G.
\newblock {\em Scheme: An Interpreter for Extended Lambda Calculus}; \hl{AI Memo
  No, MIT, Artificial Intelligence Laboratory:  1975.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).

@@

\bibitem[Meijer(2009)]{rx-observable}
Meijer, E.
\newblock Democratizing The Cloud With The .NET Reactive Framework Rx.
\newblock In Proceedings of the Internaional Softare Development Conference, Vancouver, BC, Canada, 16--24 May 2009 %MDPI: 
;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} accessed on 6 August 2025. %MDPI:

@@

\bibitem[Vogel and Springer(2022)]{progressive-2022}
Vogel, L.; Springer, T.
\newblock User Acceptance of Modified Web Page Loading Based on Progressive
  Streaming.
\newblock In Proceedings of the International Conference on Web Engineering, 2022; %MDPI: please add location and date for conference
  Springer:  Berlin/Heidelberg, Germany, %We added the location of publisher. Please confirm
 2022; pp. 391--405.

@@

\bibitem[Farago et~al.(2007)Farago, Williams, Walsh, Whyte, Goel, and
  Fung]{scroll2007}
Farago, J.; Williams, H.; Walsh, J.; Whyte, N.; Goel, K.; Fung, P.
\newblock Object Search UI and Dragging Object Results\hl{,  ~23 2007.} %MDPI: please confirm whether can be deleted
\newblock US Patent Applications 11/353,787, \hl{2007}%MDPI: please add day and month
.

\bibitem[Schiller(2007)]{schiller2007progressive}
Schiller, S.
\newblock Progressive Loading.
\newblock US Patent Applications 11/364,992, \hl{2007.} %MDPI: please add day and month

@@

\bibitem[Burke(2013)]{burke2013restful}
Burke, B.
\newblock {\em RESTful Java with JAX-RS 2.0: Designing and Developing
  Distributed Web Services}; \hl{O'Reilly Media, Inc.: 2013.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).

@@

\bibitem[Beronić et~al.(2022)Beronić, Modrić, Mihaljević, and
  Radovan]{9803765}
Beronić, D.; Modrić, L.; Mihaljević, B.; Radovan, A.
\newblock Comparison of Structured Concurrency Constructs in Java and Kotlin---Virtual Threads and Coroutines.
\newblock In Proceedings of the 2022 45th Jubilee International Convention on
  Information, Communication and Electronic Technology (MIPRO), \hl{Opatija, Croatia, 23--27 May 2022}%MDPI: Please confirm added date and location for conference
; pp. 1466--1471.
\newblock {\url{https://doi.org/10.23919/MIPRO55190.2022.9803765}}.

@@

\bibitem[Maurer and Wolfthal(2015)]{netty}
Maurer, N.; Wolfthal, M.
\newblock {\em Netty in Action}; \hl{Manning: 2015.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).

\bibitem[{\v{S}}imatovi{\'c} et~al.(2025){\v{S}}imatovi{\'c}, Markulin,
  Beroni{\'c}, and Mihaljevi{\'c}]{vsimatovic2025evaluating}
{\v{S}}imatovi{\'c}, M.; Markulin, H.; Beroni{\'c}, D.; Mihaljevi{\'c}, B.
\newblock Evaluating Memory Management and Garbage Collection Algorithms with
  Virtual Threads in High-Concurrency Java Applications.
\newblock In Proceedings of the 48th ICT and Electronics Convention MIPRO 2025. Rijeka: Croatian Society for Information, Communication and Electronic Technology–MIPRO, Opatija, Croatia, 2--6 June 2025. %MDPI: Please confirm added conference date and location.

>>>>

\begin{table}[H]
  \small
    \caption{
    Comparison of web template technologies in the Java ecosystem.
$^{(1)}$ Enforced via annotation processor that checks templates against typed interfaces at compile time.
$^{(2)}$ Compile-time expression validation available via \texttt{@CheckedTemplate} and build-time metadata in Quarkus.
$^{(3)}$ \texttt{Publisher} of reactive streams~\cite{ReactiveStreams}. Limited to a single model per web template.
$^{(4)}$ \texttt{kotlin.sequences.Sequence}
$^{(5)}$ {Non-safety} for HTML attributes.}

@@

  % Subfloat 1
\begin{minipage}[c]{0.33\linewidth}
  \centering
  \begin{lstlisting}[
  language=java,
  basicstyle=\scriptsize\ttfamily,
  numbers=none
]
html();
 head();
  title();text("JT");end();
 end();
 body();
  p();
   text("Hi JATL");
  end();
 end();
end();
  \end{lstlisting}
    \caption*{\centering(\textbf{a}) Function sequence}
  \end{minipage}
  \hfil
  % Subfloat 2
  \begin{minipage}[c]{0.23\linewidth}
    \centering
    \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html(
 head(
  title("ST")
 ),
 body(
  p("Hi ScalaTags")
 )
);
  \end{lstlisting}
    \caption*{\centering(\textbf{b}) Nested function}
  \end{minipage}
  \hfil
  % Subfloat 3
  \begin{minipage}[c]{0.32\linewidth}
    \centering
    \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html()
 .head()
  .title().text("HF").__()
 .__()
 .body()
  .p()
   .text("Hi HtmlFlow")
  .__()
 .__()
.__();
  \end{lstlisting}
    \caption*{\centering(\textbf{c}) Method chaining}
  \end{minipage}
\vspace*{6pt}
  \caption{Utilizing DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
  \label{fig:dsl-idioms}
\end{figure}

@@

All benchmarks were conducted on a GitHub-hosted virtual machine under GitHub
Actions (GitHub, Inc., San Francisco, CA, USA), running Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
(Advanced Micro Devices, Inc., Santa Clara, CA, USA)
operating at 3.24 GHz, configured with two CPU cores and two
threads per core, and 7.8 GB of available RAM. The system utilizes a 75 GB SSD
with ext4 file system, achieving 1.5 GB/s write throughput in basic I/O tests.
Network connectivity is provided through a 1500 MTU Ethernet interface. All
tests were conducted on OpenJDK 21 (Corretto build) with the G1 garbage
collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
     -Xmx16g}. Each test was executed five times per configuration to ensure result
reliability.

@@

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
WebFlux using the \texttt{Presentation} class, showing scalability results as
the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
across different rendering approaches.}
    \label{fig:presentations-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:presentations-webflux-jmeter} show that when
using blocking template engines with a separate coroutine dispatcher, the
engines are unable to scale effectively beyond four concurrent users. In contrast,
\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4487
requests per second. When blocking approaches are executed in the context of
virtual threads---thus enabling non-blocking I/O---the engines scale up to 64
concurrent users, with JStachio using virtual threads reaching 3514 requests
per second. The Thymeleaf implementation using the reactive View Resolver
driver scales up to 32 concurrent users, achieving a lower maximum throughput
of 2559 requests per second.

@@

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
MVC using the \texttt{Presentation} class, showing scalability results as
the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
across different rendering approaches.}
\label{fig:presentations-springmvc-jmeter}
\end{figure}

@@

\vspace{-3pt}
\begin{figure}[H]
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Throughput in requests per second (\texttt{req/sec}) for Quarkus
      using the \texttt{Presentation} class, showing scalability results as the
      number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
      across different rendering approaches.}
\label{fig:presentations-quarkus-jmeter}
\end{figure}

@@

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
WebFlux using the \texttt{Stock} class, showing scalability results as
the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
across different rendering approaches.}
\label{fig:stocks-webflux-jmeter}
\end{figure}

@@

\begin{figure}[H]
\vspace{-3pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
MVC using the \texttt{Stock} class, showing scalability results as
the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
across different rendering approaches.}\label{fig:stocks-springmvc-jmeter}
\end{figure}
\vspace{-6pt}
\begin{figure}[H]

     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{Throughput in requests per second (\texttt{req/sec}) for Quarkus
      using the \texttt{Stock} class, showing scalability results as the
      number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
      across different rendering approaches.}
\label{fig:stocks-quarkus-jmeter}
\end{figure}

@@

\bibitem[Carvalho(2025)]{wise2024pssr}
Carvalho, F.M.
\newblock Progressive Server-Side Rendering with Suspendable Web Templates.
\newblock In \emph{Web Information Systems Engineering---WISE
  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; Springer
: Singapore,  2025; pp. 458--473.

@@

\bibitem[Carvalho(2017)]{htmlflow}
Carvalho, F.M.
\newblock HtmlFlow Java DSL to Write Typesafe HTML.
\newblock Technical Report, 2017. Available online: \url{https://htmlflow.org/} accessed on 6 August 2025.

@@

\bibitem[Carvalho et~al.(2020)Carvalho, Duarte, and Gouesse]{carvalho2020}
Carvalho, F.M.; Duarte, L.; Gouesse, J.  
\newblock Text Web Templates Considered Harmful.  
\newblock In {\em Lecture Notes in Business Information Processing}; Springer: Cham, Switzerland, 2020; pp. 69--95.

@@

\bibitem[Sussman and Steele(1975)]{scheme}
Sussman, G.; Steele, G.  
\newblock {\em Scheme: An Interpreter for Extended Lambda Calculus};  
AI Memo No. 349, MIT Artificial Intelligence Laboratory: Cambridge, MA, USA, 1975.

@@

\bibitem[Meijer(2009)]{rx-observable}
Meijer, E.
\newblock Democratizing The Cloud With The .NET Reactive Framework Rx.
\newblock In Proceedings of the Internaional Softare Development Conference, Vancouver, BC, Canada, 16--24 May 2009 
;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} accessed on 6 August 2025.

@@

\bibitem[Vogel and Springer(2022)]{progressive-2022}
Vogel, L.; Springer, T.
\newblock User Acceptance of Modified Web Page Loading Based on Progressive
  Streaming.
\newblock In Proceedings of the International Conference on Web Engineering, 2022;
  Springer:  Berlin/Heidelberg, Germany, 
 2022; pp. 391--405.

@@

\bibitem[Farago et~al.(2007)Farago, Williams, Walsh, Whyte, Goel, and
  Fung]{scroll2007}
Farago, J.; Williams, H.; Walsh, J.; Whyte, N.; Goel, K.; Fung, P.
\newblock Object Search UI and Dragging Object Results, February 14, 2007. 
\newblock US Patent Applications 11/353,787, 2007.

\bibitem[Schiller(2007)]{schiller2007progressive}
Schiller, S.
\newblock Progressive Loading.
\newblock US Patent Applications 11/364,992, February 26, 2007.

@@

\bibitem[Burke(2013)]{burke2013restful}
Burke, B.
\newblock {\em RESTful Java with JAX-RS 2.0: Designing and Developing
  Distributed Web Services}; O'Reilly Media, Inc.: Sebastopol, CA, USA, 2013.

@@

\bibitem[Beronić et~al.(2022)Beronić, Modrić, Mihaljević, and
  Radovan]{9803765}
Beronić, D.; Modrić, L.; Mihaljević, B.; Radovan, A.
\newblock Comparison of Structured Concurrency Constructs in Java and Kotlin---Virtual Threads and Coroutines.
\newblock In Proceedings of the 2022 45th Jubilee International Convention on
  Information, Communication and Electronic Technology (MIPRO), Opatija, Croatia, 23--27 May 2022
; pp. 1466--1471.
\newblock {\url{https://doi.org/10.23919/MIPRO55190.2022.9803765}}.

@@

\bibitem[Maurer and Wolfthal(2015)]{netty}
Maurer, N.; Wolfthal, M.
\newblock {\em Netty in Action}; Manning Publications: Shelter Island, NY, USA, 2015.

\bibitem[{\v{S}}imatovi{\'c} et~al.(2025){\v{S}}imatovi{\'c}, Markulin,
  Beroni{\'c}, and Mihaljevi{\'c}]{vsimatovic2025evaluating}
{\v{S}}imatovi{\'c}, M.; Markulin, H.; Beroni{\'c}, D.; Mihaljevi{\'c}, B.
\newblock Evaluating Memory Management and Garbage Collection Algorithms with
  Virtual Threads in High-Concurrency Java Applications.
\newblock In Proceedings of the 48th ICT and Electronics Convention MIPRO 2025. Rijeka: Croatian Society for Information, Communication and Electronic Technology–MIPRO, Opatija, Croatia, 2--6 June 2025.
